Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-13 paper/ms.tex
Major changes to results
The git diff on this was weird so I just committed everything at once :/

\subsection{Abundance Distribution}\label{ssec:abundplane}
In Figure~\ref{fig:fig1}, briefly discussed in Section~\ref{sec:intro}, we show the abundance distribution of the Milky Way as well as two of our idealized merger simulations in the upper panels. A number of our idealized simulations exhibit either a bimodal or unimodal abundance distribution, and so we have selected two representative examples. The simulation labeled bimodal had orbital parameters of $R_0=142\,\kpc$, $V_0=116\,\kms$, and $\eta=0.4$. The unimodal simulation is identical, except that $R_0=129\,\kpc$. The bimodal and unimodal labels are of the outcome of the simulation, and do not reflect any particular choice made in their setup.

There are, of course, differences between the bimodal simulation and the Milky Way. First, the scaling in \MgFe{} is different -- in the simulation, the low-$\alpha$ sequence lies at $\sim0.2$, while in the Milky Way it is at about $\MgFe\sim0$. Second, in the Milky Way the high-$\alpha$ sequence neatly joins the low-$\alpha$ sequence at $\FeH\sim0$, while in the simulation the two actually diverge more at higher \FeH{}. Finally, at high \FeH{} the distribution bends upwards while in the simulation it bends downwards.

In the lower panels of Figure~\ref{fig:fig1}, we show the distributions of \MgFe{} at different fixed \FeH{}. The (blue, orange, red, green) lines show the \MgFe{} distribution at a \FeH{} of ($-0.5$, $-0.25$, $0$, $0.25$), in bins of width $0.05\,\dex$. The distributions of \MgFe{} are offset (but not rescaled) so that they do not overlap. Here, the bimodality seen in the Milky Way is quite striking at lower metallicities. The peaks are well-separated, by $\sim0.2\,\dex$. In the bimodal simulation, the distribution is still clearly bimodal, but the peaks are less well-separated, by $\sim0.1\,\dex$. In the unimodal simulation, there is a hint of some structure at $\FeH \gtrsim 0.25$, but there is not a strong multimodal structure.
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{figure1.pdf}
  \caption{The abundance bimodality seen in the Milky Way can be reproduced in some idealized merger simulations. In the upper panels, we show the distribution of stars in the \MgFe{}-\FeH{} plane. The lower panels show the distribution of \MgFe{} at a fixed \FeH{} bin of width $0.05\,\dex$. The colors indicate the fixed \FeH values, which are $-0.5$, $-0.25$, $0$, and $0.25$. The left column shows the observed distribution in the Milky Way from ASPCAP DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G}, while the right two columns show two idealized merger simulations. The idealized merger simulations are nearly identical, except that in the bimodal simulation the satellite has a starting \edit1{velocity of $142\,\kms$}, while in the unimodal simulation it has a starting \edit1{velocity of $116\,\kms$}. The labels ``unimodal'' and ``bimodal'' are of the \textit{outcome} of the simulation, and do not reflect a particular choice in the setup. The Milky Way (left column) exhibits a strong bimodal distribution of \MgFe{} at various \FeH{}. The idealized merger simulation marked as bimodal (center column) also exhibits a bimodal distribution of \MgFe{}, though the structure is not as strongly defined. The idealized merger simulation marked as unimodal (right column) exhibits only weak structure, if any at all.}
  \label{fig:fig1}
\end{figure*}

\subsection{Build up of the Abundance Plane}\label{ssec:abundplane_build}
Next, we examine the build up of the abundance plane. First, in Figure~\ref{fig:before_after}, we show the abundance plane at various points along the merger. We have divided the history of the merger into ``before'' ($t < 1.5\,\Gyr$), ``during'' ($1.5\,\Gyr < t < 2.5\,\Gyr$), and ``after'' ($t > 2.5\,\Gyr$). These times have been chosen somewhat by the orbital history and somewhat by visually inspecting the simulation, and so these deliniations are, to an extent, arbitrary. Nonetheless, it is clear that the high-$\alpha$ sequence forms before and during the merger (middle panels), and that the low-$\alpha$ sequence forms after the merger (right panel). We show a dashed line given by $-0.1\times\FeH + 0.31$, which is a crude way of separating the high- and low-$\alpha$ sequence.

It is informative to examine the star formation history (SFH) split into the high- and low-$\alpha$ sequences at a fixed \FeH{}, shown in Figure~\ref{fig:before_after_sfh}. The high- and low-$\alpha$ sequences (blue and orange) are defined as the star particles lying above and below the dashed line in Figure~\ref{fig:before_after}, respectively. We show this only in a narrow bin of width $0.1\,\dex$ at $\FeH\sim0$. We chose this value of \FeH{} because here the valley of the bimodality is deepest (see Figure~\ref{fig:fig1}). One can see a very clear separation in time between the high- and low-$\alpha$ sequences, with the high-$\alpha$ sequence forming before the low-$\alpha$ sequence.\footnote{There is a small amount of low-$\alpha$ formation at the very end of the high-$\alpha$ sequence, but this can be understood as simply an inadequacy in our simple way of separating the two sequences. Alternatively, one could assert that the separation in time is the real delineation between the two sequences, and that the boundary in the abundance plane has only mild contamination.} Furthermore, there is a brief period of $\sim300\,\Myr$ between the two sequences where no stars have formed.

It can also be shown that the exact timing of this quiescent period is metallicity-dependent. We show the same SFH in Figure~\ref{fig:before_after_sfh_by_iron}, but at varying bins in \FeH{} of $-0.5$, $-0.25$, and $0$ (blue, orange, and red, respectively). For clarity, we do not separate the SFHs by the high- and low-$\alpha$ sequences, but have verified that the same demarcation arises as in Figure~\ref{fig:before_after_sfh}. Here, we can see that at $\FeH=-0.5$ the quiescent period begins $\sim250\,\Myr$ earlier than at $\FeH=0$.

\subsection{\alphaFe{} Behavior Associated with Quiescence}
We now examine more closely the natural suspicion that the quiescent phase in Figure~\ref{fig:before_after_sfh} is responsible for the bimodal structure in Figure~\ref{fig:before_after}. In Figure~\ref{fig:SFR_alpha}, we show the SFR (blue lines) of each of the bimodal and unimodal galaxies (left and right panels, respectively) as well as the median \MgFe{} of the gas (orange lines) in a bin centered on $\FeH{}=-0.1$ with width $0.05\,\dex$. The SFR is shown for the entire galaxy ($r<15\,\kpc$) while the median \MgFe{} is shown for $2\,\kpc<r<5\,\kpc$.\footnote{The solar neighborhood-like selection of stars (discussed in Section~\ref{ssec:solarneigh}) mostly selects stars at $2\,\kpc<r<5\,\kpc$. We choose to plot the median \MgFe{} of the gas at these radii, since it is roughly that gas which is forming stars that ends up in our sample. On the other hand, we choose to show the SFR for the entire galaxy, as the enrichment from that star formation affects the entire galaxy at this epoch.}

The merger appears to induce a starburst, where the SFR rises from $\sim3.5\,\Msunyr$ at $\sim1\Gyr$ to a peak of $\sim12\,\Msunyr$ at $\sim2\,\Gyr$. This starburst is then followed by a quiescent period, with a minimum SFR of $\sim1\,\Msunyr$ and a duration of $\sim500\,\Myr$. The dip in SFR is not as pronounced as it was when we plotted the SFH at a particular iron (Figure~\ref{fig:before_after_sfh}). This quiescent period is associated with a drop in \MgFe{} from $\sim0.35$ to $\sim0.2\,\dex$ in only $\sim250\,\Myr$. Once the SFR recovers $\sim500\,\Myr$ later, \MgFe{} again rises, but only to $\sim0.28\,\dex$.

This behavior is absent in the unimodal simulation (right panel) -- while there is a starburst, it does not display a period of suppressed star formation, and there is no associated drop in \MgFe{}. It does still have a period of enhanced star formation, but this is actually coincident with an enhancement in \MgFe{}. This sequence of events is further emphasized in Figure~\ref{fig:alpha_vs_tform}, which shows a transparent scatter plot of \MgFe{} as a function of formation time. In the bimodal simulation, there is a clear gap in \MgFe{} during the period of quiescence followed by the brief formation of $\alpha$-deficient stars. In the unimodal simulation, on the other hand, there is no gap and instead a brief formation of $\alpha$-enhanced stars.

\subsection{Cause of Quiescence}\label{ssec:cause_qui}
Why there is a quiescent period in the bimodal simulation but not the unimodal simulation probably has to do with some transient, stochastic property of the gas in and around the galaxy. In some states, the AGN is better able to quench the galaxy than in others. Verifying this further is beyond the scope of this work, but we show some evidence of the association in Appendix~\ref{app:cause_qui}.

This outcome suggests that a merger alone is not sufficient in order to generate a bimodality. Rather, the merger must induce a quenching phase. Whether this occurs depends both on the properties of the merger but also on the state of the central galaxy itself and whether it can launch AGN-driven winds that result in significant gas removal. Our work would argue for some stochasticity in whether a particular merger leads to such quenching.

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{before_after.pdf}
  \caption{\textbf{The high-$\alpha$ sequence forms before the merger, the low-$\alpha$ sequence forms after the merger.} This plot shows the sequence of events leading to the build-up of the low- and high-$\alpha$ sequences for our fiducial bimodal simulation. We have separated the high- and low-$\alpha$ sequences by a dashed line at $-0.1\FeH + 0.31$, which was chosen by eye to lie in the trough. The left panel shows all star particles in our solar neighborhood cut. The middle left panel shows the star particles that form before the merger ($\tform < 1.5\,\Gyr$), which form a weak sequence of star particles at the lowest \FeH and highest \MgFe. The middle right panel shows the star particles that form during the starburst ($1.5\,\Gyr < \tform < 2.5\,\Gyr$). These star particles form the portion of the high-$\alpha$ sequence closest to the trough, and the density of star particless is higher than those that form before. The right panel shows the star particles which form after the merger ($\tform > 2.5\,\Gyr$). These star particles form almost entirely below the trough.}
  \label{fig:before_after}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{before_after_sfh.pdf}
  \caption{\textbf{At fixed metallicity, the low- and high-$\alpha$ sequence are cleanly separated in time by an intervening quiescent period.} Here we show the star formation history of the high- and low-$\alpha$ sequences. The blue (low-$\alpha$) and orange (high-$\alpha$) histograms correspond to the dashed line cut made in the \MgFe{}-\FeH{} plane in Figure~\ref{fig:before_after}, at a fixed metallicity of $\FeH=0$ with width $0.05\,\dex$. One can see that there is a nearly perfect separation in time between the formation of the high-$\alpha$ and low-$\alpha$ sequences by a quiescent period lasting $\sim300\,\Myr$. There is a small amount of contamination at the end of the high-$\alpha$ sequence which can be attributed to the inadequacy of the linear cut model of the two sequences.}
  \label{fig:before_after_sfh}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{before_after_sfh_by_iron.pdf}
  \caption{\textbf{The timing of the quiescent period which divides the high- and low-$\alpha$ sequences is metallicity dependent.} The star formation history of stars at various fixed metallicities ($\FeH=0$, $-0.25$, and $-0.5\,\textrm{dex}$) in bins of width $0.05\,\dex$. As in Figure~\ref{fig:before_after_sfh}, there is a quiescent period period which separates the formation of the low- and high-$\alpha$ sequence. However, the timing of the quiescent period is metallicity-dependent. It occurs earlier at lower metallicities, by $\sim250\,\Myr$ for the metallicities shown in this plot.}
  \label{fig:before_after_sfh_by_iron}
\end{figure}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{SFR_alpha.pdf}
  \caption{\textbf{A global suppression of star formation is associcated with a decrease in \MgFe{}, which is seen in the bimodal simulation but not in the unimodal simulation.} Here, we show both the SFR of the central galaxy ($r<15\,\kpc$) and the median \MgFe{} for gas at $2\,\kpc<r<5\,\kpc$ at a fixed \FeH{} bin centered on $0$ with width $0.05\,\dex$. The left panel shows the bimodal simulation while the right panel shows the unimodal simulation. A vertical dashed line at $2\,\Gyr$ in each plot indicates the approximate time of the merger, but before complete coalescence. In the left panel, a shaded region from $2.5\,\Gyr$ to $2.75\,\Gyr$ indicates the approximate period of quiescence shown in Figure~\ref{fig:before_after_sfh}. This suppression of star formation is associated with a sudden drop in the median \MgFe{} of the gas. Neither the suppression of star formation nor the drop in \MgFe{} are seen in the unimodal simulation.}
  \label{fig:SFR_alpha}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{alpha_vs_tform.pdf}
  \caption{\textbf{In the bimodal simulation, a gap in stellar ages is briefly followed by the formation of $\alpha$-deficient star particles.} The formation times and \MgFe{} values of star particles in the bimodal (blue) and unimodal (orange) simulations. In the unimodal simulation, no gap is present and there is briefly the formation of $\alpha$-enhanced star particles at the same time. The \MgFe{} values of the unimodal simulation have been offset by $-0.3$ for clarity.}
  \label{fig:alpha_vs_tform}
\end{figure}

>>>>

\subsection{Abundance Distribution}\label{ssec:abundplane}
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{figure1.pdf}
  \caption{The abundance bimodality seen in the Milky Way can be reproduced in some idealized merger simulations. In the upper panels, we show the distribution of stars in the \MgFe{}-\FeH{} plane. The lower panels show the distribution of \MgFe{} at a fixed \FeH{} bin of width $0.05\,\dex$. The colors indicate the fixed \FeH values, which are $-0.5$, $-0.25$, $0$, and $0.25$. The left column shows the observed distribution in the Milky Way from ASPCAP DR17 \citep[][J.A.~Holtzman et al., in preparation]{2016AJ....151..144G}, while the right two columns show two idealized merger simulations. The idealized merger simulations are nearly identical, except that in the bimodal simulation the satellite has a starting \edit1{velocity of $142\,\kms$}, while in the unimodal simulation it has a starting \edit1{velocity of $116\,\kms$}. The labels ``unimodal'' and ``bimodal'' are of the \textit{outcome} of the simulation, and do not reflect a particular choice in the setup. The Milky Way (left column) exhibits a strong bimodal distribution of \MgFe{} at various \FeH{}. The idealized merger simulation marked as bimodal (center column) also exhibits a bimodal distribution of \MgFe{}, though the structure is not as strongly defined. The idealized merger simulation marked as unimodal (right column) exhibits only weak structure, if any at all.}
  \label{fig:fig1}
\end{figure*}

In Figure~\ref{fig:fig1}, we show the abundance distribution of the Milky Way as well as two of our idealized merger simulations in the upper panels. A number of our idealized simulations exhibit either a bimodal or unimodal abundance distribution, and so we have selected two representative examples \edit1{as discussed in Section~\ref{ssec:orbit_setup}.} The bimodal and unimodal labels are of the outcome of the simulation, and do not reflect any particular choice made in their setup.

There are, of course, differences between the bimodal simulation and the Milky Way. First, the scaling in \MgFe{} is different -- in the simulation, the low-$\alpha$ sequence lies at $\sim0.2$, while in the Milky Way it is at about $\MgFe\sim0$. Second, in the Milky Way the high-$\alpha$ sequence neatly joins the low-$\alpha$ sequence at $\FeH\sim0$, while in the simulation the two actually diverge more at higher \FeH{}.

In the lower panels of Figure~\ref{fig:fig1}, we show the distributions of \MgFe{} at different fixed \FeH{}. The (blue, orange, red, green) lines show the \MgFe{} distribution at a \FeH{} of ($-0.5$, $-0.25$, $0$, $0.25$), in bins of width $0.05\,\dex$. The distributions of \MgFe{} are offset (but not rescaled) so that they do not overlap. Here, the bimodality seen in the Milky Way is quite striking at lower metallicities. The peaks are well-separated, by $\sim0.2\,\dex$. In the bimodal simulation, the distribution is still clearly bimodal, but the peaks are less well-separated, by $\sim0.1\,\dex$. In the unimodal simulation, there is a hint of some structure at $\FeH > 0.25$, but there is not a strong multimodal structure.

\subsection{Build up of the Abundance Plane}\label{ssec:abundplane_build}
Next, we examine the build up of the abundance plane. \edit1{In the left panel of Figure~\ref{fig:before_after}, we show the metallicity-dependent star formation rate of the bimodal simulation using star particles in the solar neighborhood at the end of the simulation within a $0.1\,\dex$ bin centered at $\FeH=0$. There is a clear dip in the SFR at $\sim2.8\,\Gyr$, which we indicate with a vertical line.}

\edit1{In the center left panel, we show the abundance plane distribution from the bimodal simulation for all stars in the solar neighborhood, replicating Figure~\ref{fig:fig1}. A dashed line at $-0.1\times\FeH+0.3$, chosen by eye, is plotted to demarcate the high- and low-$\alpha$ sequences. In the center right and right panels, we show the distribution of star particles which form before and after the dip in the metallicity-dependent SFR at $t=2.8\,\Gyr$, respectively. The vast majority of the high-$\alpha$ sequence forms before the dip, while most of the low-$\alpha$ sequence forms afterward.}

\edit1{This sequence of build up is markedly different in the unimodal simulation, which we show in Figure~\ref{fig:before_after_uni}. Here, there is no clear dip in the metallicity dependent SFR in the left panel. We still place a vertical line similar to the one in Figure~\ref{fig:before_after}, except now it is chosen to be at $t=2.3\,\Gyr$, which is $\sim300\,\Myr$ after the second pericentric passage. This is where the gap appears in the bimodal simulation (see Figure~\ref{fig:orbit}). We see that stars which form before and after this point in the simulation overlap considerably in the abundance plane.}

\edit1{The timing and duration of the metallicity-dependent SFR dip is not the same for all \FeH{}. In the upper panel of Figure~\ref{fig:before_after_sfh_by_iron}, we show the SFR at metallicities of $-0.5$, $-0.25$, $0$, and $0.25$ in blue, orange, red, and green, respectively. We mark the location of the dip in the red $\FeH=0$ SFR with a vertical line at $t=2.8\,\Gyr$, as in Figure~\ref{fig:before_after}. In the orange SFR at $\FeH=-0.25$, which displayed a prominent bimodality in Figure~\ref{fig:fig1}, there is a similar dip. However, it occurs about $250\,\Myr$ earlier. The $\FeH=0$ and $\FeH=-0.25$ dips have widths of about $500$ and $250\,\Myr$, respectively. In the blue $\FeH=-0.25$ SFR, which does not display a prominent bimodality, there is no dip separating two periods of sustained SF. In the green $\FeH=0.25$ SFR, a small amount of SF occurs before an extended $\sim1\,\Gyr$ dip, leading to a bimodality with a weak although well-separated high-$\alpha$ mode.}

\edit1{In the lower panel of Figure~\ref{fig:before_after_sfh_by_iron}, we show corresponding curves for the unimodal simulation. A vertical line is shown at $2.3\,\Gyr$, as described earlier. Here, we see that at no \FeH{} is there a prominent dip.}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{before_after.pdf}
  \caption{\edit1{The buildup of the abundance plane in the bimodal simulation. The left panel shows the metallicity-dependent star formation rate (SFR) for star particles in the solar neighborhood at the end of the simulation, selected within a $0.1\,\dex$ bin centered at $\FeH=0$. A clear dip in this SFR occurs at $t \sim 2.8\,\Gyr$, marked by the vertical line. The center-left panel shows the abundance plane distribution for all stars in the solar neighborhood, with a dashed line at $-0.1\times\FeH+0.3$ (chosen by eye) demarcating the high- and low-$\alpha$ sequences. The center-right and right panels show the abundance plane distributions for stars formed before and after the SFR dip, respectively. The majority of the high-$\alpha$ sequence forms before the dip, while most of the low-$\alpha$ sequence forms afterward.}}
  \label{fig:before_after}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{before_after_uni.pdf}
  \caption{\edit1{The buildup of the abundance plane in the unimodal simulation, similar to Figure~\ref{fig:before_after}. The left panel shows the metallicity-dependent star formation rate (SFR) for stars in the solar neighborhood, with no clear dip. A vertical line at $t=2.3\,\Gyr$, corresponding to the orbital stage of the gap in the bimodal simulation (see text), is included for comparison. The center left shows the abundance plane for all stars. The center-right and right panels show the abundance plane for stars formed before and after $t=2.3\,\Gyr$, respectively. Unlike the bimodal case, stars formed before and after this time overlap considerably in the abundance plane, indicating the absence of a distinct separation between high- and low-$\alpha$ sequences.}}
  \label{fig:before_after_uni}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{before_after_sfh_by_iron.pdf}
  \caption{\edit1{Metallicity-dependent star formation histories (SFH) for the bimodal (top) and unimodal (bottom) simulations. The SFH at different metallicities ($\FeH=-0.5, -0.25, 0.0, 0.25$) is shown in blue, orange, red, and green, respectively. In the top panel, the vertical line at $t=2.8\,\Gyr$ marks the SFR dip at $\FeH=0$, corresponding to the gap seen in Figure~\ref{fig:before_after}. A similar dip is present in the orange $\FeH=-0.25$ SFR, but it occurs $\sim250\,\Myr$ earlier, with a width of about $250\,\Myr$. The $\FeH=-0.5$ (blue) SFR lacks a distinct dip, while the $\FeH=0.25$ (green) SFR features an extended $\sim1\,\Gyr$ dip after a short period of SF, leading to a weak but well-separated high-$\alpha$ mode. The bottom panel shows the corresponding SFHs for the unimodal simulation, with a vertical line at $t=2.3\,\Gyr$. There are no strong dips at any metallicity, consistent with the absence of a well-defined bimodality.}}
  \label{fig:before_after_sfh_by_iron}
\end{figure}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{alpha_vs_tform_obserr.pdf}
  \caption{\edit1{In the bimodal simulation, a gap appears when plotting stars in a $0.05\,\dex$ bin at $\FeH=0$ on the \MgFe{}-$t_{\textrm{form}}$ plane (blue, left panel). In the unimodal simulation there is no such gap (orange, left panel). In the middle panel, we show the same data but assuming errors in \FeH{}, \MgFe{}, and $t_{\textrm{form}}$ of $0.0075\,\dex$, $0.012\,\dex$, and $1\,\Gyr$, respectively. The gap in the bimodal simulation appears weakly as two populations overlapping in age. In the right panel we show stars in the Milky Way from APOKASC3 in $0.2\,\dex$ bins at $\FeH=-0.5$ (blue) and $0$ (orange). The low-metallicity bin is where the Milky Way bimodality is strongest while there is no bimodality at the solar-metallicty bin (see Figure~\ref{fig:fig1}). Offsets have been added to \MgFe{} values for clarity: $-0.3$ in the unimodal points in the left and middle panels, and $+0.1$ and $-0.1$ in the low/solar-metallicity bins, respectively, in the right panel.}}
  \label{fig:alpha_vs_tform}
\end{figure*}

\subsection{Comparison to Observations}\label{ssec:compare_obs}
\edit1{A plot of \MgFe{} vs age or formation time is a useful way to further demonstrate the formation scenario of the bimodality, as well as making comparison to observations. We show this for the simulation data in the left panel of Figure~\ref{fig:alpha_vs_tform} in bins of width $0.05\,\dex$ centered at $\FeH=0$. The bimodal simulation (blue, upper data) displays a gap in the ages at $\sim2.8\,\Gyr$, coinciding with the gap in Figure~\ref{fig:before_after}. On the other hand, the unimodal simulation (orange, lower data) displays no such age gap.}

\edit1{The center panel shows these simulation data points convolved with realistic observational errors. We assume errors in \FeH{}, \MgFe{}, and $t_{\textrm{form}}$ of $0.0075\,\dex$, $0.012\,\dex$, and $1\,\Gyr$, respectively.\footnote{The \FeH{} error impacts which star particles lie in the \FeH{} selection.} The \FeH{} and \MgFe{} errors are characteristic of the errors provided in the APOGEE dataset, and the $1\,\Gyr$ comes from assuming the typical $12.5\%$ uncertainty from APOKASC-3 at an age of $8\,\Gyr$. In the bimodal simulation (blue, upper data), separate populations can be seen that overlap significantly in $t_{\textrm{form}}$. There is not a strong separation in the unimodal case (orange, lower data).}

\edit1{The right panel shows observational data from APOKASC-3. We show in the blue, upper data a $0.2\,\dex$ bin centered at $\FeH=-0.5$ and in the orange, lower data centered at $\FeH=0$. In the Milky Way, there is a bimodal and unimodal \MgFe{} distribution at these metallicities, respectively. At $\FeH=-0.5$, one could weakly argue that there are separate populations as in the simulation. However, the large uncertainties at the relevant ages ($\sim1\,\Gyr$ at ages of $\sim8\,\Gyr$) and the small sample size (125 and 1175 at $\FeH=-0.5$ and $0$, respectively) prevents a definitive statement from being made. Larger sample sizes and more precise age estimates in the future would clarify the connection between simulation and observation, although achieving age uncertainities of $<10\%$ for such old stars is very challenging \citep[e.g.][]{2010ARA&A..48..581S}.}

\edit1{There is an additional complication in the data coming from the presence of young, $\alpha$-rich stars. These have been argued to be old stars with misclassified astroseismic ages due to binary mass transfer \citep[and references therein]{2023A&A...671A..21J}, though with some appearing to be genuinely young \citep[and references therein]{2024arXiv241002962L}, with a range of explanations given \citep[e.g.][]{2015A&A...576L..12C,2021MNRAS.508.4484J,2023arXiv231105815S}. It is unclear which of the $\alpha$-rich stars faithfully reflect the mean ISM chemistry at their inferred age, and so it is not obvious if any detailed conclusions can be drawn from this comparison.}

\subsection{Full Simulation Suite}\label{ssec:full_sim_suite}
\edit1{We next expand our discussion to the full simulation suite by studying the conditional 1D \MgFe{} distribution in Figure~\ref{fig:all_hist}. We plot the distribution of \MgFe{} for all stars with \FeH{} lying in a bin of width $0.1\,\dex$ centered at $0\,\dex$. Each distribution corresponds to a choice in the three orbital variables $R_0$, $V_0$, and $\eta$. We rank the simulations in alphabetical order from least to most bimodal\footnote{There is one more simulation in the suite than letters in the English alphabet, so the simulation with the highest bimodality score is labeled aa.}, as determined by the bimodality score $\mathcal{B}$ introduced in Section~\ref{ssec:bim_score}. The list of simulations in the orbital grid along with the associated orbital parameters and bimodality scores is given in Table~\ref{tab:my-table}.}

\edit1{Simulations from r onwards, which have $\mathcal{B}>2.25$, are considered bimodal. This demarcation was chosen by eye using Figure~\ref{fig:all_hist}. For these simulations, we plot the trough \MgFe{} (the location of the minimum between the two maxima) as a vertical line. This value was determined by taking the location of the minimum of the distribution between $0.25$ and $0.35\,\dex$, and is also given in Table~\ref{tab:my-table}.}

\edit1{Simulations from r to aa have clear bimodalities. The trough \MgFe{} is roughly consistent between simulations, appearing between $\sim0.25$ and $0.3\,\dex$. For simulations marked as unimodal, some appear to have two populations but which are not distinct enough to form a clear bimodality (e.g., m, n, and o).}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{hist.pdf}
  \caption{\edit1{The conditional 1D \MgFe{} distribution for all stars with \FeH{} in a bin of width $0.1\,\dex$ centered at $0\,\dex$. Each panel corresponds to a different set of orbital parameters ($R_0$, $V_0$, and $\eta$) and is labeled alphabetically from least to most bimodal, with the labels defined in Table~\ref{tab:my-table}. The final simulation, which has the highest bimodality score $\mathcal{B}$, is labeled ``aa.'' Simulations with $\mathcal{B}>2.25$ (starting from simulation r) exhibit clear bimodalities, with the trough \MgFe{} (minimum between the two peaks) marked. This trough generally falls between $0.25$ and $0.3\,\dex$. Simulations with lower bimodality scores appear unimodal, though some (e.g., m, n, and o) show hints of a secondary population without forming a distinct bimodal structure.}}
  \label{fig:all_hist}
\end{figure*}

\begin{table}[]
\centering
\begin{tabular}{cccccc}
$\textrm{letter}$ & $R_0$ & $V_0$ & $\eta$ & $\mathcal{B}$ & $\textrm{trough}\,\MgFe{}$ \\ \hline
a                 & $129$ & $129$ & $0.4$  & $0.29$        &                            \\
b                 & $129$ & $116$ & $0.4$  & $0.35$        &                            \\
c                 & $116$ & $116$ & $0.4$  & $0.77$        &                            \\
d                 & $142$ & $116$ & $0.5$  & $0.79$        &                            \\
e                 & $142$ & $129$ & $0.6$  & $0.89$        &                            \\
f                 & $116$ & $129$ & $0.4$  & $1.12$        &                            \\
g                 & $116$ & $116$ & $0.5$  & $1.25$        &                            \\
h                 & $116$ & $142$ & $0.4$  & $1.34$        &                            \\
i                 & $142$ & $142$ & $0.5$  & $1.43$        &                            \\
j                 & $129$ & $142$ & $0.5$  & $1.46$        &                            \\
k                 & $142$ & $129$ & $0.4$  & $1.62$        &                            \\
l                 & $142$ & $142$ & $0.6$  & $1.75$        &                            \\
m                 & $116$ & $129$ & $0.5$  & $1.82$        &                            \\
n                 & $129$ & $129$ & $0.5$  & $1.96$        &                            \\
o                 & $129$ & $116$ & $0.5$  & $2.04$        &                            \\
p                 & $116$ & $116$ & $0.6$  & $2.16$        &                            \\
q                 & $116$ & $129$ & $0.6$  & $2.24$        &                            \\ \hline
r                 & $142$ & $129$ & $0.5$  & $2.26$        & $0.283$                    \\
s                 & $142$ & $116$ & $0.4$  & $2.52$        & $0.313$                    \\
t                 & $116$ & $142$ & $0.6$  & $2.59$        & $0.258$                    \\
u                 & $116$ & $142$ & $0.5$  & $2.62$        & $0.293$                    \\
v                 & $142$ & $116$ & $0.6$  & $2.65$        & $0.273$                    \\
w                 & $129$ & $142$ & $0.6$  & $2.66$        & $0.288$                    \\
x                 & $142$ & $142$ & $0.4$  & $2.70$        & $0.283$                    \\
y                 & $129$ & $116$ & $0.6$  & $2.94$        & $0.303$                    \\
z                 & $129$ & $129$ & $0.6$  & $3.12$        & $0.278$                    \\
aa                & $129$ & $142$ & $0.4$  & $3.34$        & $0.298$                   
\end{tabular}
\caption{\edit1{All simulations in the orbital grid (as defined by $R_0$, $V_0$, and $\eta$) ordered by their bimodality score $\mathcal{B}$. Each simulation is given an identifying letter. For simulations marked as bimodal ($\mathcal{B}>2.25$), we also list the trough \MgFe{}, or location of the minimum of the \MgFe{} distribution.}}
\label{tab:my-table}
\end{table}

\edit1{We study the formation of the 1D distributions in Figure~\ref{fig:all_hist} through a scatter plot of \MgFe{} vs formation time of star particles in Figure~\ref{fig:all_scatter}, similar to Figure~\ref{fig:alpha_vs_tform}. The order and colors are identical as in Figure~\ref{fig:all_hist}, and we use the same \FeH{} selection. We only plot a random subsample of 350 stars, and points are plotted with a transparency of $\alpha=0.5$ so that the perceived density is slightly higher in cases of overlap. An offset of $-0.2$ and $-0.4$ are given to the second and third (orange and red) simulations in each panel. For simulations marked as bimodal (r through aa), we also plot a horizontal line at the location of the trough \MgFe{}.}

\edit1{Most of the bimodal simulations (r through aa) have a gap in the distribution at around the time of the merger ($\sim2$-$3\,\Gyr$, depending on the orbital parameters). Before this gap, star formation occurs above the trough while after it occurs below the trough. However, there are two exceptions: simulation~t does not have a full gap (though the density does decrease), and simulation~w is irregular, with star formation simultaneously occuring above and below the trough.}

\edit1{In the unimodal simulations (a through q), there does appear to be gaps (d, f, g, h, and o). However, in these cases, there is not a large amount of star formation occuring before the gap, and so the high-$\alpha$ sequence is underemphasized and does not lead to a significant bimodality.}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{scatter.pdf}
  \caption{Scatter plot of \MgFe{} versus formation time for star particles in each simulation, corresponding to the 1D distributions shown in Figure~\ref{fig:all_hist}. Each panel represents a different set of orbital parameters ($R_0$, $V_0$, and $\eta$), arranged from least to most bimodal (as defined by the bimodal score $\mathcal{B}$. The colors and order match Figure~\ref{fig:all_hist}, with a random subsample of 350 stars plotted per simulation. Transparency ($\alpha=0.5$) is used to highlight regions of higher density. The second and third (orange and red) simulations in each panel are offset by $-0.2$ and $-0.4\,\dex$ for clarity, respectively. For bimodal simulations (r through aa), the trough \MgFe{} (minimum between the two peaks) is indicated with a horizontal line. In most bimodal cases, a gap in the distribution emerges at approximately the merger time ($\sim2$--$3,\Gyr$, depending on orbital parameters), with older stars forming at higher \MgFe{} and younger stars at lower \MgFe{}. Notably, simulation~t lacks a clear gap, and simulation~w exhibits irregular behavior with star formation occurring both above and below the trough. Among unimodal simulations (a through q), some exhibit apparent gaps (e.g., d, f, g, h, and o), but these do not result in strong bimodalities due to the low number of high-\MgFe{} stars forming before the gap.}
  \label{fig:all_scatter}
\end{figure*}
